Quantifying the locality of operators that translate Stokes $Q$ \& $U$ to scalars $E$ \& $B$

### 2017-10-26 articles/formalism/article.tex
made a few folder in articles named formalism. Updated abstract. Monor modifications to notebook making the modified radial kernel plots

\title{Local real space operators for translating Stokes Q/U to E/B maps on the sphere }
\author[]{Aditya Rotti and}
\author[]{Kevin Huffenberger}
\affiliation[]{Department of Physics, Florida State University, Keen Physics Building, 77 Chieftan Way, Tallahassee, Florida, U.S.A.}
\emailAdd{adityarotti@gmail.com}
\emailAdd{khuffenberger@fsu.edu}

\abstract{
\noindent We derive real space operators to translate Stokes Q/U to coordinate independent even and odd parity E/B fields on the sphere and demonstrate their numerical equivalence to the harmonic space methods. In process we also derive real space operators which allow direct decomposition of measured Q/U into Q/U fields corresponding to E-modes and B-modes respectively. These operators are derived on the sphere and account for the curvature effects. The analytical derivations reveal a functional form for the radial kernel, which quantifies the non-local nature of the E and B fields. 
%and their dependence on the band limit of the measurement. 
These real space operators allow the computation of E and B field in each pixel by integrating in a small region around it, and hence it can be easily parallelized and is not memory intensive. A major drawback of these real space operators is that they cannot reliably construct modes on large angular scales $\ell<50$ with ease. Finally, this study motivates the construction of more locally defined E and B maps, by making desired alterations to the radial filter. We demonstrate that these altered E mode and B mode constructions do not significantly distort spectral information on small angular scales. %Such locally constructed E and B maps may help mitigate the foreground leakage issue across the two modes.
}

@@

\include{./sections/introduction}
\include{./sections/real_space_formalism}
\include{./sections/kernel_discussion}
\include{./sections/generalizedeb}
\include{./sections/numerical_implementation}
\include{./sections/application_353ghz}
\include{./sections/discussion}
\include{./sections/appendix}

>>>>

\title{Quantifying the locality of operators that translate Stokes $Q$ \& $U$ to scalars $E$ \& $B$}
\title{Real space operators for Stokes $Q$ \& $U$ to scalars $E$ \& $B$ translation}
\author[]{Aditya Rotti and}
\author[]{Kevin Huffenberger}
\affiliation[]{Department of Physics, Florida State University, Keen Physics Building, 77 Chieftan Way, Tallahassee, Florida, U.S.A.}
\emailAdd{adityarotti@gmail.com}
\emailAdd{khuffenberger@fsu.edu}

\abstract{
\noindent  We derive fully sky real space operators to translate Stokes Q \& U parameters to scalars $E$ \& $B$ and vice versa. We explicitly show that these local real space operator are fully characterized by the spin-0 $Y_{\ell 2}$ spherical harmonic functions. These real space operators make transparent the association of radial and tangential patterns of polarization with the E modes and that of  the clockwise and anti-clockwise pinwheel patterns of polarization with B-modes. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidate these derivations. Using this new notation also allows us to derive real space operators which decompose the measured Stokes parameters into those corresponding to E-modes and B-modes respectively, without ever evaluating the scalar fields themselves. We use these analytical derivations us to quantify the non-local nature of the relation between the two representations of the polarization field. We present a prescription for generalizing these operators, which allow the non-local behavior of these operators to be a tunable parameter. We derive constraints on these generalized operators to ensure reliable recovery of the standard $E$ and $B$ fields and the corresponding power spectra. This is part 1 of a series of papers where we explore these operators.}

@@

\include{./sections/introduction}
\include{./sections/real_space_formalism}
\include{./sections/kernel_discussion}
\include{./sections/generalizedeb}
\include{./sections/discussion}
\include{./sections/appendix}

### 2017-10-26 articles/formalism/sections/generalizedeb.tex
made a few folder in articles named formalism. Updated abstract. Monor modifications to notebook making the modified radial kernel plots

The radial part of these generalized convolution kernels is now given by the following expressions,
%
\begin{subequations}
\beqry
G_{QU \rightarrow EB}(\beta) &=& G(\beta) = \sum _{\ell=\ell_{\rm min}} ^{\ell_{\rm max}} g_{\ell}\frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell + 2)!}} P_{\ell}^2(\cos{\beta}) \, \label{eq:mod_rad_forward} \\
G_{EB \rightarrow QU}(\beta) &=& G^{-1}(\beta) = \sum _{\ell=\ell_{\rm min}} ^{\ell_{\rm max}} g_{\ell}^{-1}\frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell + 2)!}} P_{\ell}^2(\cos{\beta}) \,,\label{eq:mod_rad_inverse}
\eeqry
\end{subequations}
%
where $g_{\ell}$ are the same multipole function as those appearing in $\tilde{\mathcal{G}}$. If one chooses different forms of $g_{\ell}$ for E and B modes then the radial function are defined accordingly. Given this general definition for the radial function $G(\beta)$, note that the default radial function $f(\beta)$ is just a special case resulting from the choice $\tilde{\mathcal{G}}=\tilde{\mathcal{I}}$ ($g_{\ell}=1$). Note that for this choice of $\tilde{\mathcal{G}}$ the inverse is trivial $\tilde{\mathcal{G}}^{-1}=\tilde{\mathcal{G}}$ and therefore $G^{-1}(\beta) = G(\beta)$.

While defining these alternate operators by modifying the radial part of the kernels, it seems more natural to make a choice on the real space function $G(\beta)$ as compared to choosing the multipole function $g_{\ell}$. Using the orthogonality property of associated Legendre polynomials it can be shown that the multipole function $g_{\ell}$ is given by the following integral once the radial function $G(\beta)$ is specified,
%
\beq
g_{\ell} = 2 \pi \sqrt{\frac{(\ell-2)!}{(\ell+2)!}} \int _{0}^{\pi} G(\beta) P_{\ell}^{2}(\cos{\beta}) d\cos{\beta} \,. \label{eq:gb2bl}
\eeq
%
Here it is important to note that the radial function has to be chosen such that $G(\beta)$ has to be such that it vanishes at $\beta=0$ and $\beta=\pi$. One way to understand this is that the associated Legendre polynomials $P_{\ell}^2$ vanish at these values of the abscissa and hence cannot be used to describe function which don't have this property.  Another way to understand this requirements is that at these locations the coordinate dependence of the Stokes parameters cannot be integrated out, since the azimuthal angle is ill defined and hence the convolution kernel needs to have vanishing contribution from these locations. \revisit{The normalization of these functions is not important, since that just defines a convention. One just needs to ensure to be consistent with the convention once a choice has been made.} \comment{find a better location for this}

In contrast,  the beam convolution function $B(\beta) \rightarrow 1$ as $\beta \rightarrow 0$ and a circularly symmetric beam's harmonic representation $b_{\ell}$ are the coefficients of expansion of $B(\beta)$ in the Legendre polynomial $P_{\ell}^0$ basis. \comment{Distinguishing E/B localization characterized by $g_{\ell}$ from E/B smoothing characterized by $b_{\ell}$.}

The discussion till now gives the impression that using the localized convolution kernels is no different from from using the default kernel and altering the spherical harmonic coefficients of expansion of the relavant fields by appropriately operating on them with the  effective beam functions $g_{\ell}$. To appreciate the difference between these two, it is important to realize that in general one can make a choice of a radial function which may not have a band limited description. In such a case these two method of evaluating the relevant fields is not identical. An example of this claim is depicted in \fig{fig:example_gbeta}.  
%
\begin{figure}[!t] 
\centering
\subfigure[]{\includegraphics[width=0.49\columnwidth]{example_gbeta.pdf}}
\subfigure[]{\includegraphics[width=0.49\columnwidth]{example_eff_bl.pdf}}
\caption{}
\label{fig:example_gbeta}
\end{figure}
%

Another important thing to realize is that the harmonic coefficients derived from default full sky operations get some contributions from different portions of sky. For instance evaluating the E and B fields in the vicinity of the poles is are prone to receiving significant contributions from strong foregrounds near the equator. Correcting the harmonic coefficients of expansion with the effective beam function does not cancel these non-local contribution. On the contrary by performing the convolution with the localized real space kernels, the regions which contribute to the local field evaluations are predetermined by the choice of the radial function.

@@

\subsection{Recovering the default E and B mode spectra}
The convolution kernels defined using these modified radial functions returns some E' and B' mode maps,
%
\beq
\bar{S}' = \bar{O}' * \bar{P}
\eeq
%
 which are not the standard $E$ and $B$ modes maps. We would now like to relate the spectra of these modified $E'$ and $B'$ mode maps to the standard spectra. Even though the $G(\beta)$ maybe such that it doesn't have an accurate band limited  description, the spectra are always evaluated to a definite band limit. Therefore the harmonic representation $g_{\ell}$ of the radial function $G(\beta)$ can be used to appropriately normalize the spectra so as to match the standard E and B mode spectra.
 Specifically it can be shown that the spectra of $E'$ and $B'$ field is related to the spectra of the spectra of the standard E and B fields via the following simple relation,
 %
 \beq
 C_{\ell}^{E',B'} =  C_{\ell}^{E,B} g_{\ell}^2\,,
 \eeq
 %
%--------------------------------------------------------
%--------------------------------------------------------

>>>>

The radial part of these generalized convolution kernels is now given by the following expressions,
%
\begin{subequations}
\beqry
G_{QU \rightarrow EB}(\beta) &=& G(\beta) = \sum _{\ell=2} ^{\ell_{\rm max}} g_{\ell}\frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell + 2)!}} P_{\ell}^2(\cos{\beta}) \, \label{eq:mod_rad_forward} \\
G_{EB \rightarrow QU}(\beta) &=& G^{-1}(\beta) = \sum _{\ell=2} ^{\ell_{\rm max}} g_{\ell}^{-1}\frac{2 \ell+1}{4 \pi} \sqrt{\frac{(\ell-2)!}{(\ell + 2)!}} P_{\ell}^2(\cos{\beta}) \,,\label{eq:mod_rad_inverse}
\eeqry
\end{subequations}
%
where $g_{\ell}$ are the same multipole function as those appearing in $\tilde{\mathcal{G}}$. If one chooses different forms of $g_{\ell}$ for E and B modes then the radial function are defined accordingly. Given this general definition for the radial function $G(\beta)$, note that the default radial function $f(\beta)$ is just a special case resulting from the choice $\tilde{\mathcal{G}}=\tilde{\mathcal{I}}$ ($g_{\ell}=1$). Note that for this choice of $\tilde{\mathcal{G}}$ the inverse is trivial $\tilde{\mathcal{G}}^{-1}=\tilde{\mathcal{G}}$ and therefore $G^{-1}(\beta) = G(\beta)$.

While defining these alternate operators by modifying the radial part of the kernels, it seems more natural to make a choice on the real space function $G(\beta)$ as compared to choosing the multipole function $g_{\ell}$. Using the orthogonality property of associated Legendre polynomials it can be shown that the multipole function $g_{\ell}$ is given by the following integral over the radial function $G(\beta)$,
%
\beq
g_{\ell} = 2 \pi \sqrt{\frac{(\ell-2)!}{(\ell+2)!}} \int _{0}^{\pi} G(\beta) P_{\ell}^{2}(\cos{\beta}) d\cos{\beta} \,. \label{eq:gb2bl}
\eeq
%
Here it is important to note that the radial function $G(\beta)$ has to be chosen such that it vanishes at $\beta=0$ and $\beta=\pi$. One way to understand this is that the associated Legendre polynomials $P_{\ell}^2 \propto \sin^2{\beta}$ vanish at these values of the abscissa and hence cannot be used to describe functions which don't have this property.  Another way to understand this requirements is that at these locations the coordinate dependence of the Stokes parameters cannot be integrated out, since the azimuthal angle is ill defined and hence the convolution kernel needs to have vanishing contribution from these locations. \revisit{The normalization of these functions is not critically important, since any choice defines a convention. It is important to ensure consistency with the convention once a choice has been made.} \comment{find a better location for this}

In contrast to the radial function $G(\beta)$ an instrumental beam function appropriately normalized has the property $B(\beta) \rightarrow 1$ as $\beta \rightarrow 0$. A circularly symmetric beam is represented in harmonic space by the multipole function $b_{\ell}$ which can be derived by evaluating the coefficients of expansion of the function $B(\beta)$ defined at the pole in the Legendre polynomial $P_{\ell}^0$ basis. The effective instrument beam is \revisit{almost} always identical for $E$ and $B$ modes. \comment{Distinguishing E/B localization characterized by $g_{\ell}$ from E/B smoothing characterized by $b_{\ell}$.}
%
\beq
B(\beta) = \sum_{\ell=0}^{\ell_{\rm max}} \frac{2 \ell+1}{4 \pi} b_{\ell} P_{\ell}^{0} (\cos{\beta})\,,
\eeq
%
\comment{A co-polarized beam which measures the Stokes parameters is not the same as a smoothing beam measuring a scalar field. Is it important to discuss the polarization beam details here ?}
%
\begin{figure}[!t] 
\centering
\subfigure[\label{fig:bl_gbeta}]{\includegraphics[width=0.32\columnwidth]{Gbeta_for_different_gl_lmax1536.pdf}}
\subfigure[]{\includegraphics[width=0.32\columnwidth]{gl_for_different_gbeta_lmax1536.pdf}}
\subfigure[\label{fig:gl_bbeta}]{\includegraphics[width=0.32\columnwidth]{Bbeta_for_gl_lmax1536.pdf}}
\caption{\textit{Left:} The vertical dashed gray line depicts the approximate pixel size $\Delta_{\rm pix} = \sqrt{\frac{4 \pi}{N_{\rm pix}}}$ of a Nside=512 Healpix map. The green line depicts the default radial kernel $f(\beta)$ defined in \eq{eq:qu2eb_gen_kernel}. The blue and orange lines depict the modified radial function resulting the beam harmonics $b_{\ell}$ corresponding to Gaussian beams with fwhm=15 \& 12 arcminutes respectively. The red curve depicts an example modified radial function: $G(\beta)=\mathcal{N} \beta^n \exp{\left[ -\left( \frac{\beta-\beta0}{\sqrt{2} \sigma} \right)^s \right]}$ with parameters set to the following values $[n=1;\, \beta_0=0 ;\, \sigma = 2\Delta_{\rm pix} ;\, s=1.5]$. The black dashed curve depicts the band limited reconstruction of the modified radial function $G(\beta)$. We intentionally have plotted $\beta G(\beta)$ to clearly depict the high $\beta$ behavior of these functions. \textit{Middle: } This figure depicts the harmonic representation of the respective radial functions as indicated by the legend. The dashed curves of the corresponding color depict the inverse of the harmonic functions. \textit{Right:} This figure depicts the beam function $B(\beta)$ evaluated from interpreting the respective harmonic functions as those corresponding to an instrument beam.}
\label{fig:example_gbeta}
\end{figure}
%
Though the real space behavior of these two function $G(\beta)$ and $B(\beta)$ has important differences, in harmonic space they play identical roles. Therefore it is possible to interpret the beam harmonic coefficients as those representing some modified radial kernel. The modified radial kernel resulting from a few example Gaussian beams are depicted in \fig{fig:bl_gbeta}.
it is possible to interpret the harmonic representation $g_{\ell}$ of the radial function $G(\beta)$ as those corresponding to some circularly symmetric instrument beam function. The beam function corresponding to the default radial kernel and the modified radial kernel $G(\beta)$ are depicted in \fig{fig:gl_bbeta}.

\comment{Move the following two paragraphs to final discuss}
\revisit{The discussion till now gives the impression that using the localized convolution kernels is no different from from using the default kernel and altering the spherical harmonic coefficients of expansion of the relavant fields by appropriately operating on them with the  effective beam functions $g_{\ell}$. To appreciate the difference between these two, it is important to realize that in general one can make a choice of a radial function which may not have a band limited description. In such a case these two method of evaluating the relevant fields is not identical. An example of this claim is depicted in \fig{fig:example_gbeta}.\\
Another important thing to realize is that the harmonic coefficients derived from default full sky operations get some contributions from different portions of sky. For instance evaluating the E and B fields in the vicinity of the poles is are prone to receiving significant contributions from strong foregrounds near the equator. Correcting the harmonic coefficients of expansion with the effective beam function does not cancel these non-local contribution. On the contrary by performing the convolution with the localized real space kernels, the regions which contribute to the local field evaluations are predetermined by the choice of the radial function.}

@@

\subsection{Recovering the default E and B mode spectra}
The convolution kernels defined using a modified radial function $G(\beta)$ returns some scalar $E'$ and $B'$ mode maps,
%
\beq
\bar{S}' = \bar{O}' * \bar{P}
\eeq
%
which are not the standard $E$ and $B$ modes maps. Since the the harmonic representation $g_{\ell}$ of the radial function $G(\beta)$ can be simply interpreted as the harmonic coefficients of some beam,  the spectra of the scalar fields $E'$ and $B'$ are related to the spectra of the spectra of the standard $E$ and $B$ fields via the following relation, 
 %
 \begin{subequations}
 \beqry
C_{\ell}^{EE,BB,EB} &= &C_{\ell}^{E'E',B'B',E'B'} /   g_{\ell}^2\,.\\
C_{\ell}^{TE,TB}  &=&  C_{\ell}^{TE',TB'} / g_{\ell}\,.
 \eeqry
 \end{subequations}
 %
The auto and cross spectra with the standard $E$ and $B$ fields can be recovered from the modified fields $E'$ and $B'$. Their accurate recovery only relies on the inverse of the harmonic functions $1/g_{\ell}$ being well behaved, which can be ensured by smartly choosing the radial function $G(\beta)$.

Unlike the instrument beam, the radial function $G(\beta)$ is an analysis choice. functions are perfectly known. Since the radial kernel is perfectly known it is in principle possible to perfectly deconvolve the same. In contrast the beam function $B{\beta}$ has to be inferred from measurement of point source and hence is not known exactly and therefore the deconvolution is only approximate.
Even though the $G(\beta)$ maybe such that it doesn't have an accurate band limited  description, the spectra are always evaluated to a definite band limit.  
%--------------------------------------------------------
%--------------------------------------------------------

### 2017-10-26 articles/formalism/sections/introduction.tex
made a few folder in articles named formalism. Updated abstract. Monor modifications to notebook making the modified radial kernel plots

\section{Introduction}

@@

\item Healpix has rings at constant latitude. So the real space convolution is only perfect for pixels near the pole ? What happens when you consider pixels around any arbitrary pixel on a Healpix map, they are probably not arranged in rings. Maybe for this reason working with upgraded map works and finer pixels allow for a better approximation of rings around arbitrary pixels ??
\end{enumerate}

>>>>

\item Healpix has rings at constant latitude. So the real space convolution is only perfect for pixels near the pole ? What happens when you consider pixels around any arbitrary pixel on a Healpix map, they are probably not arranged in rings. Maybe for this reason working with upgraded map works and finer pixels allow for a better approximation of rings around arbitrary pixels ??

\item The Zaldariagga radial function $1/\theta^2$ cannot be correct, since it does not vanish at $\beta=\pi$, while it is defined to be zero at $\beta=0$. I guess since its in the flat sky limit, the $\beta=\pi$ issue is not relevant.
\end{enumerate}

\textbf{To do list:}
\begin{enumerate}
\item Check whether the local Healpix works for primordial B-mode spectrum.
\item Check the generation of local kernels by using a an effective beam in harmonic space. pl2 transformed
\item How do the  coefficients of expansion for some radially compact kernel, derived from pl2 transforms differ from that of a pl0 transform ? 
\end{enumerate}

\newpage
\section{Introduction}

@@
